---
title: "Robinson G.K.(1991), That BLUP Is a Good Thing: The Estimation of Random Effects"
subtitle: "Paper Summary"
author: "Peter von Rohr"
date: "`r Sys.Date()`"
output:
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_html: default
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

# Disclaimer
This is a summary of the paper [@Robinson1991]. This aims at finding an easy enough^[Lecture Notes by [@Goetz2014] mention that none of the available derivations of BLUP is suitably easy to be presented in an Agronomy Masters Class] derivation of BLUP, such that it can be presented in class. 


# Introduction
BLUP stands for _Best Linear Unbiased Prediction_ and is the de-facto standard method for predicting or estimating^[In German, only the translated term of estimation is used.] breeding values. It is a method of estimating random effects^[Therefore, the model is called a linear mixed effects model.]. 

The context of BLUP is the linear model

\begin{equation}
y = X\beta + Zu + e
\label{eq:LinMixedModel}
\end{equation}

\begin{tabular}{llp{10cm}}
where  &  &  \\
       &  $y$  &  vector of $n$ observable random variables \\
       &  $\beta$  &  vector of $p$ unknown parameters having fixed size (fixed effects) \\
       &  $X$, $Z$ &  known design matrices \\
       &  $u$, $e$ &  vectors of $q$ and $n$ unobservable random variables (random effects)
\end{tabular}

{The first and second moments of the random effects are given as}

$$E(u) = 0 \text{ and } E(e) = 0$$ and 

$$Var\left[\begin{array}{c} u \\ e \end{array} \right] 
  = \left[ \begin{array}{cc}  
    G  &  0 \\
    0  &  R
  \end{array}\right] * \sigma^2 
$$  

where $G$ and $R$ are positive definite matrices and $\sigma^2$ is a known constant.

In general, it is assumed that variance-covariance structures are known^[The topic of variance components estimation is addressed later], except for the constant factor $\sigma^2$. 

BLUP estimates of the realisations of the random variables $u$ are 

* _linear_ in the sense that they are linear functions of the data $y$. 
* _unbiased_ in the sense that the average estimate is equal to the average of the quantity that is estimated^[shorter we can say that $E(u) = E(\hat{u})$].
* _best_ in the sense that they have minimum error in the class of all linear unbiased estimators and 
* _predictors_ to distinguish them from estimators.

{Discussion about difference between predictor and estimator is ignored here}

Mathematically, the BLUP estimators $\hat{\beta}$ of $\beta$ and $\hat{u}$ of $u$ are defined as the solution to the following simultaneous equations, given by [@Henderson1950]. 

\begin{eqnarray}
X^TR^{-1}X\hat{\beta} + X^TR^{-1}Z\hat{u}  &=&  X^TR^{-1}y \nonumber \\
Z^TR^{-1}X\hat{\beta} + (Z^TR^{-1}Z + G^{-1}) \hat{u}  &=&  Z^TR^{-1}y
\label{eq:MixedModelEq}
\end{eqnarray}

The above given equations are called _mixed model equations_ (MME) and $\hat{\beta}$ and $\hat{u}$ are called mixed model solutions. We note that as $G^{-1}$ tends to $0$, the MME tend towards the generalized least-squares (GLS) equations for estimating $\beta$ and $u$, when these components are regarded as fixed effects.

[@Henderson1975] showed that, provided $X$ has full column rank $p$, the variance-covariance matrix of estimation errors is given by 

\begin{equation}
E \left(
    \left[ 
      \begin{array}{c}
      \beta - \hat{\beta}\\
      u - \hat{u}
      \end{array}
    \right]
\left[ 
      \begin{array}{c}
      \beta - \hat{\beta}\\
      u - \hat{u}
      \end{array}
    \right]^T
\right)
=
\left[
  \begin{array}{cc}
  X^TR^{-1}X  &  X^TR^{-1}Z \\
  Z^TR^{-1}X  &  Z^TR^{-1}Z + G^{-1}
  \end{array}
\right]^{-1} \sigma^2
\label{eq:CovEstError}
\end{equation}

The BLUP estimates generally differ from the generalized least squares estimates that would be obtained if $u$ were regarded to be fixed. This is illustrated by the following example.

## Example 
A simple example of model (\ref{eq:LinMixedModel}) is that of first lactation yields of dairy cows with sire additive genetic merit^[also known as breeding values] being treated as random effects ($u$) and herds being treated as fixed effects ($\beta$).  The matrix $R\sigma^2$ is the variance-covariance matrix of the vector ($e$) which corresponds to the departures from a model where observations are entirely explained by sire effects and herd effects^[shorter, we just write $var(e) = R\sigma^2$]. We are assuming that the matrix $G$ is known as a multiple of the identity matrix $I$. For our example, we use $G = 0.1*I$. 

Suppose the following dataset is given

```{r, echo=FALSE, results='asis'}
tbl_sire_data <- dplyr::data_frame(Herd = c(rep(1,2), rep(2,3),rep(3,4)),
                                   Sire = c("A", "D","B","D","D","C", "C","D","D"),
                                   Yield = c(110,100,110,100,100,110,110,100,100))
knitr::kable(tbl_sire_data, align = "c", booktabs = TRUE)
```

The entities in equation (\ref{eq:LinMixedModel}) are 







```{r bib, include=FALSE}
vec_bref <- c(bibentry(
  bibtype = "Article",
  title = "That BLUP Is a Good Thing: The Estimation of Random Effects",
  author = as.person("G.K. Robinson"),
  year = 1991,
  journal = "Statistical Science",
  volume = 6,
  number = 1,
  page = "15-51",
  key = "Robinson1991"),
  bibentry(
    bibtype = "Unpublished",
    title   = "Quantitative Genetik und Zuchtplanung",
    author  = as.person("K-U. Götz [aut]"),
    year    = "2014",
    note    = "Bayerische Landesanstalt für Landwirtschaft, Institut für Tierzucht",
    key     = "Goetz2014"
  ),
  bibentry(
    bibtype = "Article",
    title = "Estimation of genetic parameters (abstract)",
    author = as.person("C.R. Henderson"),
    year = 1950,
    journal = "Ann. Math. Statist.",
    volume = 21,
    page = "309-310",
    key = "Henderson1950"
  ),
  bibentry(
    bibtype = "Article",
    title = "Best linear unbiased estimation and prediction under a selection model",
    author = as.person("C.R. Henderson"),
    year = 1975,
    journal = "Biometrics",
    volume = 31,
    page = "423-447",
    key = "Henderson1975"
  ))
# write entries in vec_bref to bib-file
sBibFile <- "skeleton.bib"
if(!file.exists(sBibFile))
  cat(paste(toBibtex(vec_bref), collapse = "\n"), "\n", file = sBibFile)

```
